### 2022-01-28 .debris/2022-01-27 15.13.53.00/Paper.tex
modified

\section{Sticky air and free surface}\label{sec:sticky}
The Earth's surface can be treated by means of either the so-called sticky-air or a true free surface method, both of them implemented in the code. In the
sticky-air method the surface is approximated with the introduction of a buoyant layer with a viscosity at least four orders of magnitude lower than the crust
\citep{Schmeling2008,Crameri2012} and the interface between lithosphere and air is defined using a chain of passive markers that are advected as the Lagrangian
markers. The correctness of the evolution of the markers chain is tested with the experiment of a 2D time-dependent Stokes sphere below a free surface and
compared with results from ASPECT\footnote{\url{https://aspect.geodynamics.org/}} \citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} (Section
\ref{sec:sphere}). In the true free surface case the top boundary is assumed stress-free and velocities are not fixed. In this case topography variations are
described by vertical deformations of the mesh that depend on the velocity field of the nodes that identify the free surface, while horizontal deformations are
not taken into account. This procedure is known as the Arbitrary Lagrangian-Eulerian (ALE) method and its implementation follows the technique described in
\citet{Thieulot2011}. However, although the implementation of a true free surface better reproduces laboratory experiment, extremely small time steps can be
necessary to maintain stability \citep{Kaus2010a,Quinquis2011,Thieulot2014}. Therefore, the stability algorithm proposed by \citet{Kaus2010a} is implemented
to avoid instabilities due to high density differences at the free surface when using too large time steps. The implementation of this algorithm are tested by
performing the experiment described by \citet{Kaus2010a} (Section \ref{sec:stab}).

>>>>

\section{Sticky air and free surface}\label{sec:sticky}
The Earth's surface can be treated by means of either the so-called sticky-air or a true free surface method, both of them implemented in the code. In the
sticky-air method the surface is approximated with the introduction of a buoyant layer with a viscosity at least four orders of magnitude lower than the crust
\citep{Schmeling2008,Crameri2012} and the interface between lithosphere and air is defined using a chain of passive markers that are advected as the Lagrangian
markers. The correctness of the evolution of the markers chain is tested with the experiment of a 2D time-dependent Stokes sphere below a free surface and
compared with results from ASPECT\footnote{\url{https://aspect.geodynamics.org/}} \citep{Kronbichler2012,Heister2017,aspect,aspectmanual} (Section
\ref{sec:sphere}). In the true free surface case the top boundary is assumed stress-free and velocities are not fixed. In this case topography variations are
described by vertical deformations of the mesh that depend on the velocity field of the nodes that identify the free surface, while horizontal deformations are
not taken into account. This procedure is known as the Arbitrary Lagrangian-Eulerian (ALE) method and its implementation follows the technique described in
\citet{Thieulot2011}. However, although the implementation of a true free surface better reproduces laboratory experiment, extremely small time steps can be
necessary to maintain stability \citep{Kaus2010a,Quinquis2011,Thieulot2014}. Therefore, the stability algorithm proposed by \citet{Kaus2010a} is implemented
to avoid instabilities due to high density differences at the free surface when using too large time steps. The implementation of this algorithm are tested by
performing the experiment described by \citet{Kaus2010a} (Section \ref{sec:stab}).

### 2022-01-28 .debris/2022-01-27 15.42.10.00/Paper.tex
modified

\subsection{Instantaneous 2D sphere}\label{sec:ist_sphere}
The domain is a unit square with gravity $\bm{g}=(0,-1)$. The fluid has constant density and viscosity ($\rho_f=1$ and $\eta_f=1$). The sphere is in the middle
of the domain with a radius R = 0.123456798 and has constant density and viscosity ($\rho_s=10^{-2}+\rho_f$ and $\eta_s=10^3 \cdot \eta_f$).
We distinguish three different types of velocity boundary conditions:
\begin{enumerate}
\item FS: free slip conditions on all sides.
\item NS: no slip conditions on all sides.
\item OT: free slip conditions on the sides and the bottom, and open boundary at the top.
\end{enumerate}
All three cases are performed for different grid resolutions, between \num{16x16} and \num{512x512} elements with 50 randomly distributed markers per element,
and different average schemes for the viscosity (harmonic, geometric and arithmetic). The velocity in the centre of the sphere $\bm{v}(0.5;0.5)$), minimum and
maximum velocities ($u_{min}$, $u_{max}$, $v_{min}$, $v_{max}$) and pressures ($p_{min}$, $p_{max}$) on the entire domain, $v_{\textrm{rms}}$ and average pressure on the
entire domain ($p_{avg}$) are compared with solutions generated by ASPECT \citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} and all results
can be found at \url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere2D} and 
\url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum_equation/Instantaneous_2D_Sphere}. Results in terms of the $v_{\textrm{rms}}$ for NS model 
are also shown in Fig. \ref{fig:inst_sphere}.

@@

\subsection{2D time-dependent Stokes sphere with deformable free surface}\label{sec:sphere}
This experiment is performed in a unit square domain with the gravity acceleration fixed to $g_y=-1$. The sphere has $\rho_s=2$ and $\eta_s=10^3$ and is
initially centred at ($0.5;0.6$) with radius $R=0.123456789$. The fluid surrounding the sphere has $\rho_f=1$ and $\eta_f=1$ and occupies the domain for
$y\leq 0.75$, while for $y>0.75$ the air has $\rho_a=0$ and $\eta_a=10^{-3}$. Velocity boundary conditions are set to free slip on all sides. The Courant
number is set to 0.25. The experiments run for 200 s using grid resolutions from \num{150x150} to \num{512x512} elements, with 25 randomly distributed markers
per element, and for different average schemes for the viscosity. The interface between the fluid and the air is tracked by means of the markers chain. Results
in terms of velocity and pressure fields and topography variations are compared with results obtained with ASPECT
\citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} and they can be found at
\url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere_fs2D} and
\url{https://github.com/aleregorda/Benchmarks/tree/main/Surface_processes/Time_dependent_sphere}. Results in terms of the $v_{\textrm{rms}}$ in case of 
an arithmetic average are also shown in Fig. \ref{fig:inst_spherefs}.

@@

The experiment is performed with two grid resolution (\num{125x40} and \num{200x80} elements), with 100 markers per element and CFL of 0.25. $v_{\textrm{rms}}$
(see Sec. \ref{sec:error}) calculated for both the simulations match well with results from \citet{vanKeken1997} and obtained with ASPECT
\citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} and ELEFANT \citep{Thieulot2014}, obtained with similar grid resolutions
(Fig. \ref{fig:thin}). In addition, a variety of simulations with different resolutions and aspect ratios are performed to verify that the sensitivity of the
initial velocity field. Results match well with results from \citet{vanKeken1997} (in the grey area) and \citet{Thieulot2014} (Fig. \ref{fig:thin_initial}). 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum%2BEnergy/Thin_layer}.

>>>>

\subsection{Instantaneous 2D sphere}\label{sec:ist_sphere}
The domain is a unit square with gravity $\bm{g}=(0,-1)$. The fluid has constant density and viscosity ($\rho_f=1$ and $\eta_f=1$). The sphere is in the middle
of the domain with a radius R = 0.123456798 and has constant density and viscosity ($\rho_s=10^{-2}+\rho_f$ and $\eta_s=10^3 \cdot \eta_f$).
We distinguish three different types of velocity boundary conditions:
\begin{enumerate}
\item FS: free slip conditions on all sides.
\item NS: no slip conditions on all sides.
\item OT: free slip conditions on the sides and the bottom, and open boundary at the top.
\end{enumerate}
All three cases are performed for different grid resolutions, between \num{16x16} and \num{512x512} elements with 50 randomly distributed markers per element,
and different average schemes for the viscosity (harmonic, geometric and arithmetic). The velocity in the centre of the sphere $\bm{v}(0.5;0.5)$), minimum and
maximum velocities ($u_{min}$, $u_{max}$, $v_{min}$, $v_{max}$) and pressures ($p_{min}$, $p_{max}$) on the entire domain, $v_{\textrm{rms}}$ and average pressure on the
entire domain ($p_{avg}$) are compared with solutions generated by ASPECT \citep{Kronbichler2012,Heister2017,aspect,aspectmanual} and all results
can be found at \url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere2D} and 
\url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum_equation/Instantaneous_2D_Sphere}. Results in terms of the $v_{\textrm{rms}}$ for NS model 
are also shown in Fig. \ref{fig:inst_sphere}.

@@

\subsection{2D time-dependent Stokes sphere with deformable free surface}\label{sec:sphere}
This experiment is performed in a unit square domain with the gravity acceleration fixed to $g_y=-1$. The sphere has $\rho_s=2$ and $\eta_s=10^3$ and is
initially centred at ($0.5;0.6$) with radius $R=0.123456789$. The fluid surrounding the sphere has $\rho_f=1$ and $\eta_f=1$ and occupies the domain for
$y\leq 0.75$, while for $y>0.75$ the air has $\rho_a=0$ and $\eta_a=10^{-3}$. Velocity boundary conditions are set to free slip on all sides. The Courant
number is set to 0.25. The experiments run for 200 s using grid resolutions from \num{150x150} to \num{512x512} elements, with 25 randomly distributed markers
per element, and for different average schemes for the viscosity. The interface between the fluid and the air is tracked by means of the markers chain. Results
in terms of velocity and pressure fields and topography variations are compared with results obtained with ASPECT
\citep{Kronbichler2012,Heister2017,aspect,aspectmanual} and they can be found at
\url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere_fs2D} and
\url{https://github.com/aleregorda/Benchmarks/tree/main/Surface_processes/Time_dependent_sphere}. Results in terms of the $v_{\textrm{rms}}$ in case of 
an arithmetic average are also shown in Fig. \ref{fig:inst_spherefs}.

@@

The experiment is performed with two grid resolution (\num{125x40} and \num{200x80} elements), with 100 markers per element and CFL of 0.25. $v_{\textrm{rms}}$
(see Sec. \ref{sec:error}) calculated for both the simulations match well with results from \citet{vanKeken1997} and obtained with ASPECT
\citep{Kronbichler2012,Heister2017,aspect,aspectmanual} and ELEFANT \citep{Thieulot2014}, obtained with similar grid resolutions
(Fig. \ref{fig:thin}). In addition, a variety of simulations with different resolutions and aspect ratios are performed to verify that the sensitivity of the
initial velocity field. Results match well with results from \citet{vanKeken1997} (in the grey area) and \citet{Thieulot2014} (Fig. \ref{fig:thin_initial}). 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum%2BEnergy/Thin_layer}.

### 2022-01-28 .debris/2022-01-27/Paper.tex
modified

\section{Erosion and sedimentation}\label{sec:erosion}
Surface processes at either the lithosphere-air interface or in correspondence of the free surface have been implemented by means of the software
Fastscape\footnote{\url{https://fastscape.org/}} \citep{Braun2013,Cordonnier2019,Yuan2019a,Yuan2019b}.FastScapeLib is a set of routines that solve the stream
power law (enriched by a sediment transport/deposition term) \citep{Yuan2019a}, hillslope diffusion and marine transport and deposition \citep{Yuan2019b}.
The partial differential equation solved by FastScapeLib is:
\begin{equation}
\frac{\partial h}{ \partial t}=U-K_f A^{m_e}S^{n_e} + \frac{G}{A}\int_A \left( U-\frac{\partial h}{\partial t} \right) dA + K_d \bm{\nabla}^2 h 
\end{equation}
where $h$ is the topography, $U$ is the uplift, $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $S$ is the slope, $m_e$ and $n_e$ are
parameters that should be chosen to constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}, $G$ is the dimensionless
deposition/transport coefficient for the enriched stream power law and $K_d$ is the effective diffusivity. Terms $$K_f A^{m_e}S^{n_e}$$ and $$K_d \bm{\nabla}^2 h$$
consider long-range processes described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020} and short-range processes are
proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}.

In case of sticky air, slope is determined by means of the markers chain, which is vertically corrected considering erosion and sedimentation rates. After the
correction, continental markers above the markers chain are transformed in air markers and air markers below the markers chain become sediments. Similarly,
slope is determined considering the nodes on the top boundary in case of true free surface and they are vertically corrected as for the markers chain. In this
case, continental markers above the top boundary are simply deleted.
%Surface processes at the lithosphere-air interface are implemented taking into account fluvial erosion (long-range processes), hillslope processes (short-range
%processes) and failure of slopes exceeding a threshold angle $\theta$ \citep{Burov1997,Tucker1999,Sacek2012,Croissant2014,Ueda2015,Beucher2020}. Erosion rates
%$\dot{E}_{lr}$ due to long-range processes can be described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{lr}=K_f A^{m_e} \bm{\nabla} h^{n_e}
%\end{equation}
%where $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $\bm{\nabla} h$ is the slope and $m_e$ and $n_e$ are parameters chosen in to
%constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}. Erosion rates $\dot{E}_{sr}$ due to short-range processes are
%implemented by assuming that transport is linearly proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{sr}=K_d \bm{\nabla}^2 h
%\end{equation}
%where $K_d$ is the effective diffusivity. In addition, the slope is modified if it still exceeds a threshold angle after long- and short-range processes
%\citep{Ueda2015}.

%Offshore sedimentation processes are implemented similarly to hillslope processes but with a depth-dependent coefficient \citep{Sacek2012,Beucher2020}.
%The sedimentation rate $\dot{S}$ is calculated as
%\begin{equation}
%\dot{S}=K_s \bm{\nabla}^2 h
%\end{equation}
%with
%\begin{equation}
%K_s=K^*_s \exp\left(\frac{h-h_{sl}}{h_0}\right)
%\end{equation}
%where $K^*_s$ is a constant, $h_{sl}$ is the sea level and $h_0$ is a characteristic length. In case of sticky air, slope is determined by means of the markers
%chain, which is vertically corrected considering erosion and sedimentation rates. After the correction, continental markers above the markers chain are
%transformed in air markers and air markers below the markers chain become sediments. Similarly, slope is determined considering the nodes on the top boundary
%in case of true free surface and they are vertically corrected as for the markers chain. In this case, continental markers above the top boundary are simply
%deleted.

>>>>

\section{Erosion and sedimentation}\label{sec:erosion}
Surface processes at the lithosphere-air interface are implemented taking into account fluvial erosion (long-range processes), hillslope processes (short-range
processes) and failure of slopes exceeding a threshold angle $\theta$ \citep{Burov1997,Tucker1999,Sacek2012,Croissant2014,Ueda2015,Beucher2020}. Erosion rates
$\dot{E}_{lr}$ due to long-range processes can be described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, in according to
\begin{equation}
\dot{E}_{lr}=K_f A^{m_e} \bm{\nabla} h^{n_e}
\end{equation}
where $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $\bm{\nabla} h$ is the slope and $m_e$ and $n_e$ are parameters chosen in to
constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}. Erosion rates $\dot{E}_{sr}$ due to short-range processes are
implemented by assuming that transport is linearly proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}, in according to
\begin{equation}
\dot{E}_{sr}=K_d \bm{\nabla}^2 h
\end{equation}
where $K_d$ is the effective diffusivity. In addition, the slope is modified if it still exceeds a threshold angle after long- and short-range processes
\citep{Ueda2015}.

Offshore sedimentation processes are implemented similarly to hillslope processes but with a depth-dependent coefficient \citep{Sacek2012,Beucher2020}.
The sedimentation rate $\dot{S}$ is calculated as
\begin{equation}
\dot{S}=K_s \bm{\nabla}^2 h
\end{equation}
with
\begin{equation}
K_s=K^*_s \exp\left(\frac{h-h_{sl}}{h_0}\right)
\end{equation}
where $K^*_s$ is a constant, $h_{sl}$ is the sea level and $h_0$ is a characteristic length. In case of sticky air, slope is determined by means of the markers
chain, which is vertically corrected considering erosion and sedimentation rates. After the correction, continental markers above the markers chain are
transformed in air markers and air markers below the markers chain become sediments. Similarly, slope is determined considering the nodes on the top boundary
in case of true free surface and they are vertically corrected as for the markers chain. In this case, continental markers above the top boundary are simply
deleted.
